A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-29 tidyindex.qmd
work on dimension reduction

When indexes are constructed from multivariate information, dimension reduction methods combine that information into a univariate series. In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):

\begin{equation}
x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
\end{equation}

where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
The most commonly used dimension reduction technique is Principal Component Analysis (PCA). It can be seen as a special case of weighting, where variables are summed up in a linear combination: $$x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$ with restrictions imposed on the weight coefficient: $\sum_{p=1}^P\lambda_p^2 = 1$. In other cases of weighting, the coefficients can be as simple as giving equal weight to each variables.

Some dimension reduction can also be formulated from domain-specific knowledge. This can be theories that describe the physics of the phenomenon being indexed or practical formulations used to extract useful features from the raw variables. For example, in the index SPEI, a difference series is calculated between precipitation and potential evapotranspiration (PET) and the validity of this formulation is backed up by climate water balance model [Thornthwaite, 1948], which describes [...]. *Add another example of remote sensing variables i.e. NDVI = (NIR - Red) / (NIR + Red)?*

@@

While suggested weights and formulas can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit the specific tasks.

>>>>

Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}

where $p = 1, 2, \cdots, P$. The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process. For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).

Another widely used approach is linear combination, which aggregates a collection of variables in a linear additive structure, expressed as: $$x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$

where $\lambda_p$ denotes the weight assigned to variable $x_p$. Most indexes uses an equal weight that sum to 1, which allows each variable to contribute equally to the index. A linear combination can also be viewed as a linear projection of multivariate information by the weight vector. Projecting data from higher to lower dimension inevitably leads to information loss. For example, two countries can receive similar index, but their components could significantly differ -- one with average scores across all components and another with extreme scores the opposite ends. The selected set of weights need to be examined to understand its effect on the index and its implications for decision-making. To do this, analysts can vary the coefficients in the linear projection to observe how the index value and countries' ranking change. These changes can also be visualized using a method called tour by generating an animation of the projections among different sets of weights.

<!-- Other weight constraint used includes $\sum_{p=1}^P\lambda_p^2 = 1$, is used by principal component analysis.  -->

@@

### 2023-07-29 tidyindex.tex
work on dimension reduction

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

When indexes are constructed from multivariate information, dimension
reduction methods combine that information into a univariate series. In
the pipeline, dimension reduction includes methods that take
multivariate inputs and output the data in a lower dimension (often
univariate):

\begin{equation}
x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
\end{equation}

where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
variable dimension from \(P\) to \(P^*\). The most commonly used
dimension reduction technique is Principal Component Analysis (PCA). It
can be seen as a special case of weighting, where variables are summed
up in a linear combination:
\[x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
with restrictions imposed on the weight coefficient:
\(\sum_{p=1}^P\lambda_p^2 = 1\). In other cases of weighting, the
coefficients can be as simple as giving equal weight to each variables.

Some dimension reduction can also be formulated from domain-specific
knowledge. This can be theories that describe the physics of the
phenomenon being indexed or practical formulations used to extract
useful features from the raw variables. For example, in the index SPEI,
a difference series is calculated between precipitation and potential
evapotranspiration (PET) and the validity of this formulation is backed
up by climate water balance model {[}Thornthwaite, 1948{]}, which
describes {[}\ldots{]}. \emph{Add another example of remote sensing
variables i.e.~NDVI = (NIR - Red) / (NIR + Red)?}

While suggested weights and formulas can indicate norms adored by
practitioners, analysts should be given the flexibility to experiment
with different combinations when constructing indexes. This could help
understand index behavior from its sensitivity to the variables and
suggest alternative weights that better suit the specific tasks.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

Dimension reduction summarises the multivariate information into
univariate, which can be denoted as:

\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}

where \(p = 1, 2, \cdots, P\). The combination can be based on
domain-specific knowledge, originated from theories describing the
underlying physical process. For example, the SPEI uses a water balance
model (\(D = P - \text{PET}\)) to calculate the difference series
(\(D\)) from precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)).

Another widely used approach is linear combination, which aggregates a
collection of variables in a linear additive structure, expressed as:
\[x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]

where \(\lambda_p\) denotes the weight assigned to variable \(x_p\).
Most indexes uses an equal weight that sum to 1, which allows each
variable to contribute equally to the index. A linear combination can
also be viewed as a linear projection of multivariate information by the
weight vector. Projecting data from higher to lower dimension inevitably
leads to information loss. For example, two countries can receive
similar index, but their components could significantly differ -- one
with average scores across all components and another with extreme
scores the opposite ends. The selected set of weights need to be
examined to understand its effect on the index and its implications for
decision-making. To do this, analysts can vary the coefficients in the
linear projection to observe how the index value and countries' ranking
change. These changes can also be visualized using a method called tour
by generating an animation of the projections among different sets of
weights.
